The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-10-09 paper.tex
Exemplo finalize: output em vez de funcao

\begin{figure}
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void STOP, RETRANSMIT, SENDACK;
par/or do
    await STOP;
with
    loop do
        par/or do
            await RETRANSMIT;
        with
            par/and do
                await 1min;
            with
                var _pkt_t buffer;
                <fill-buffer-info>
                _send_enqueue(&buffer);
                await SENDACK;
            end
        end
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.41\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=11]
<...>
    var _pkt_t buffer;
    <fill-buffer-info>
    finalize
        _send_enqueue(&buffer);
    with
        _send_cancel(&buffer);
    end
    await SENDACK;
<...>
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{The unsafe network protocol in the left, which does not compile, is 
extended with a finalization clause in the right to handle abortion properly.
%\caption{ Parallel compositions can describe complex state machines.
\label{lst.fin}
}
\end{figure}

To illustrate threats related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, which does not compile in \CEU.
It describes the state machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (ln. 1) 
represent the external interface of the protocol.
%
The protocol has to transmit a packet every minute, unless it receives a 
\code{RETRANSMIT} request to immediately re-transmit it, or a \code{STOP} 
request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (ln. 3);
the other periodically transmits the packet (ln. 5--18).
%
The periodic transmission is a loop that starts two other trails (ln. 6--17):
one handles the immediate retransmission request (ln. 7);
the other transmits the packet and waits for a confirmation (ln. 9--16).
%
%
The actual transmission (ln. 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping (ln. 10), in accordance with the 
specification.
Note that the call to \code{\_send\_enqueue}%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \C libraries explicit and are required in \CEU.
}
(ln. 14) is \emph{asynchronous}, handing to the radio driver a pointer to the 
lexically-scoped packet (ln. 12).
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SENDACK} to acknowledge completion 
(ln. 15).
%
At any time, the client may request a retransmission (ln. 7), which terminates 
the \code{par/or} (ln. 6), aborts the ongoing transmission (ln. 14, if not 
idle), and restarts the loop (ln. 5).
%
The client may also request to stop the whole protocol at any time (ln. 3).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
\CEU tracks the interaction of \code{par/or} compositions with local variables 
and stateful \C functions (e.g., device drivers) in order to preserve safe 
abortion of trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
accompany the stateful \C call.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

The $mem(id)$ primitive represents all accesses, assignments, and \C function 
calls that affect a memory location identified by $id$.
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%In accordance with the zero-delay hypothesis of \CEU, $mem$ expressions are 
%considered to be atomic and instantaneous
%
The special notation $nop$ is used to represent an innocuous $mem$ expression 
(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
unused identifier).
%
Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
an identifier is either a variable or an event.

>>>>

\begin{figure}
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void STOP, RETRANSMIT, SEND_ACK;
output _pkt_t* SEND_ENQUEUE, SEND_CANCEL;
par/or do
    await STOP;
with
    loop do
        par/or do
            await RETRANSMIT;
        with
            par/and do
                await 1min;
            with
                var _pkt_t buffer;
                <fill-buffer-info>
                emit SEND_ENQUEUE => &buffer;
                await SEND_ACK;
            end
        end
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em,firstnumber=12]
<...>
    var _pkt_t buffer;
    <fill-buffer-info>
    finalize
        emit SEND_ENQUEUE => &buffer;
    with
        emit SEND_CANCEL => &buffer;
    end
    await SEND_ACK;
<...>
\end{lstlisting}
\end{minipage}
%\rule{8.5cm}{0.37pt}
\caption{The unsafe network protocol in the left, which does not compile, is 
extended with a finalization clause in the right to handle abortion properly.
%\caption{ Parallel compositions can describe complex state machines.
\label{lst.fin}
}
\end{figure}

To illustrate threats related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, which does not compile in \CEU.
It describes the state machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input and output events represent the external interface of the protocol 
(ln. 1--2).
%
The protocol has to transmit a packet every minute (with \code{SEND\_ENQUEUE}), 
unless it receives a \code{RETRANSMIT} request to immediately re-transmit it, 
or a \code{STOP} request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (ln. 4);
the other periodically transmits the packet (ln. 6--19).
%
The periodic transmission is a loop that starts two other trails (ln. 7--18):
one handles the immediate retransmission request (ln. 8);
the other transmits the packet and waits for a confirmation (ln. 10--17).
%
%
The actual transmission (ln. 13--16) is enclosed with a \code{par/and} that 
takes at least one minute before looping (ln. 11), in accordance with the 
specification.
Note that the \code{emit SEND\_ENQUEUE} (ln. 15) is \emph{asynchronous}, 
handing to the radio driver a pointer to the lexically-scoped packet (ln. 13).
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SEND\_ACK} to acknowledge completion 
(ln. 16).
%
At any time, the client may request a retransmission (ln. 8), which terminates 
the \code{par/or} (ln. 7), aborts the ongoing transmission (ln. 15, if not 
idle), and restarts the loop (ln. 6).
%
The client may also request to stop the whole protocol at any time (ln. 4).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
\CEU tracks the interaction of \code{par/or} compositions with local variables 
and stateful output events and \C calls in order to preserve safe abortion of 
trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
accompany the output request.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

The $mem(id)$ primitive represents all accesses, assignments, \C function 
calls, and output events that affect a memory location identified by $id$.
As the challenging parts of \CEU reside on its control structures, we are not 
concerned here with a precise semantics for side effects, but only with their 
occurrences in programs.
%In accordance with the zero-delay hypothesis of \CEU, $mem$ expressions are 
%considered to be atomic and instantaneous
%
The special notation $nop$ is used to represent an innocuous $mem$ expression 
(it can be thought as a synonym for $mem(\epsilon)$, where $\epsilon$ is an 
unused identifier).
%
Note that $mem$ and $await$/$emit$ expressions do not share identifiers, i.e., 
an identifier is either a variable or an event.
